Making the Best of a Bad Situation

### 2022-03-21 paper.tex
tweaks

Much of the time, this burden can be mitigated by automation, such as when the data structure being generated
is composed entirely of structures that already have generators, or $P$ is described as an inductive
relation \cite{gggir}. However, the case of inferring generators for \emph{numerical} properties---those involving constriants over numbers--- is not well
handled by prior work. Numerical properties appear regularly in
systems programming tasks \cite{???}, where structured data described by inductive invariants is rare, inputs regularly
take the form of pointers and integers, and many function preconditions amount to bounds-checking or ensuring
that certain fields in different structures are related in some way.

@@

\section{Generator Learning with Bandits Algorithms}
\label{sec:bandits}
With our bag of generators in hand, we now need to find the best one. The
approach we take will be inspired by the Multi-Armed Bandits
\cite{gittins1979bandit} problem from the theory of reinforcement learning. In short, the multi armed
bandits problem describes a situation where an algorithm is repeatedly presented
a fixed set of choices. Each choice gives a different (random) reward, and the
goal of the game is to maximize the total reward. Solving this optimization problem
online is too difficult in an adversarial setting, so algorithms for the multi-armed
bandits problem aim to instead minimize \emph{regret}: how much worse their total reward
was than the best possible total reward in hindsight. To achieve this, bandits algorithms
learn which actions have historically given more reward, and play those more frequently.
In our setting, the ``choices" are our
generator candidates, and the rewards are given by success or failure of a
generator to yield a value. Under this analogy, an algorithm for the Multi-Armed
Bandits problem will let us learn which generators give the best results
\textit{while simultaneously} generating a stream of valid inputs for the
function.

The algorithm we will use is called UCB1 \cite{auer2002finite}. While more sophisticated
bandits algorithms exist, this one is sufficient for our purposes. When given a list of generators
$g_1,\dots g_k$ and a number of rounds $T$ to run for, UCB1 runs the generators $g_i$ in a way that attempts
to learn which generator succeds the most frequently, while also attempting to not waste time by running
failing generators too much. In the end, UCB1 emits the $m \leq T$ successfully-generated values over the course of its run.
The main upshot is that the ratio success ratio $\frac{m}{T}$ of this derived generator should, in the limit,
be no worse than the \emph{best} generator the algorithm was given. 
% UCB1 achieves a
% regret bound of $O(\sqrt{KT\log K})$. While algorithms are known which achieve
% $O(\sqrt{KT})$ regret, UCB1 is simple to implement and understand, and performs
% more than well enough for our setting. In \autoref{alg:ucb1}, we present the
% UCB1 algorithm, modified for use in property-based testing. The algorithm takes
% generator candidates $g_1,\dots g_k$, the property $P$, a number of rounds to
% run for $T$, and returns a list of generated values satisfying $P$.
\begin{restatable}{thm}{regretthm}
Fix a property $P$ and generators $g_1,\dots,g_K$, where $g_i$ succeeds with probability at least $p_i$. Then,
\vspace{-2px}
\[
\lim_{T \to \infty} \mathbb{E}\left[\frac{1}{T}\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|\right] \geq \max_i p_i
\]
\end{restatable}
% Intuitively, this theorem justifies our intuition that the ucb1 algorithm
% ``finds the best generator". As the number of rounds gets large, the ratio of
% the number of valid inputs generated by \texttt{ucb1} to the number of rounds is
% no worse than the success probability of the best generator. 
In other words, with high probability
% \footnote{One could do an
% analysis using Chebyshev's inequality to bound the difference between the
% empirical frequency of the sequence and its mean.}
, the stream of values
emitted by an ``infinite run" of UCB1 acts like a $(\max_i p_i)$-good
generator for $P$.

@@

I have implemented the generator inference algorithm described in this abstract,
as well as a deductive program verifier (a la Dafny or Frama-C \cite{??}) for the
\texttt{IMP} language which uses these generators as a verification backend.
The code is available \href{https://github.com/jdublu10/triple-testing}{here}.

% Because the generator inference algorithm implemented by the prototype is the
% first of its kind, there are no existing tools we can compare it to in a
% benchmark. Instead, we simply present our results as a table
% (\autoref{tab:data}) of data about runs of the algorithm on different inputs. We
% leave it to the reader to decide if these results are impressive.

In \autoref{tab:data}, we show the results for some trials of our algorithm. In
each trial, we run the generration algorithm on the specified constraint five
times for $T = 1000$ iterations. 
% $K$ is the average number of unique scripts
% discovered by the random DFS, Rejected is the average number of failed generator
% samples that happen during the runs, and Unique is the average number of unique
% input sets generated. In all cases, the generation takes less than 1ms of wall
% clock time. Integers are generated in the range $[-10000,10000]$ instead of the full integer
% range.

In \autoref{tab:data-selected}, we present selected results: a larger table
is available in the Appendix. The largest issue is the performance
differences between the runs on the proposition $0 \leq x \leq y \leq 100$ with
and without explicitly including the implicand $x \leq 100$.
The ``best" generator for the proposition without it samples an
arbitrary $x \geq 0$, which is very unlikely to leave room for some $y$
satisfying $x \leq y \leq 100$. Including the added constraint ensures that we leave space
with our initial choice of $x$.

@@

\begin{table}[h]
\label{tab:data-selected}
\centering
\scalebox{0.8}{\parbox{\linewidth}{%
\begin{tabular}{|l|l|l|l|l|}
\hline
Constraint & $K$ & $R$ & $U$ \\ \hline
    %  $0 \leq x \leq y \leq 100 \wedge x \leq 100$ & $2$ & $11$ & $831$ \\ \hline
     $x = y + z \wedge x \geq y \wedge x,y,z \geq 0$      &   $4$  &    $48.6$      &    $951.4$    \\ \hline
     $0 \leq x \leq y \leq 100$ & $2$ & $993.4$ & $6.6$ \\ \hline
     $0 \leq x \leq y \leq 100 \wedge x \leq 100$ & $2$ & $11$ & $831$ \\ \hline
     $x - y \leq 5 \wedge x - y \leq 10 \wedge y - z \leq 2$ & $3.8$ & $26.2$ & $831$ \\ \hline
\end{tabular}
\caption{Selected Results. $K$ = average number of unique scripts, $R$ = number of failed draws, $U$ = number of unique values generated}
}}
\end{table}

@@

\begin{center}
\begin{table}[h]
\label{tab:data-full}
\centering
\scalebox{0.8}{\parbox{\linewidth}{%
\begin{tabular}{|l|l|l|l|l|}
\hline
Constraint & $K$ & $R$ & $U$ \\ \hline
     $0 \leq x \leq y \leq 100 \wedge x \leq 100$ & $2$ & $11$ & $831$ \\ \hline
     $x = y + z \wedge x,y,z \geq 0$      &   $5$   &    $506$      &     $494$   \\ \hline
     $x = y + z \wedge x \geq y \wedge x,y,z \geq 0$      &   $4$  &    $48.6$      &    $951.4$    \\ \hline
     $0 \leq x \leq y $  &  $2$   &  $24.6$   &       $975.4$       \\ \hline
     $0 \leq x \leq y \leq 100$ & $2$ & $993.4$ & $6.6$ \\ \hline
     $0 \leq x \leq y \leq 100 \wedge x \leq 100$ & $2$ & $11$ & $831$ \\ \hline
     $x - y \leq 5 \wedge x - y \leq 10 \wedge y - z \leq 2$ & $3.8$ & $26.2$ & $831$ \\ \hline
\end{tabular}
\caption{Full Results}
}}
\end{table}
\end{center}
\noindent\rule{0.5\textwidth}{1pt}

>>>>

Much of the time, this burden can be mitigated by automation, such as when the data structure being generated
is composed entirely of structures that already have generators, or $P$ is described as an inductive
relation \cite{gggir}. However, the case of inferring generators for \emph{numerical} properties---those involving constriants over numbers--- is not well
handled by prior work. Numerical properties appear regularly in
systems programming tasks \cite{ironfleet} where structured data described by inductive invariants is rare, inputs regularly
take the form of pointers and integers, and many function preconditions amount to bounds-checking or ensuring
that certain fields in different structures are related in some way.

@@

\section{Generator Learning with Bandits Algorithms}
\label{sec:bandits}
With our bag of generators in hand, we now need to find the best one. The
approach we take will be inspired by the Multi-Armed Bandits
\cite{gittins1979bandit} problem from the theory of reinforcement learning. In short, the multi armed
bandits problem describes a game where an algorithm is repeatedly presented
a fixed set of choices. Each choice gives a different (random) reward, and the
goal of the game is to maximize the total reward. Solving this optimization problem
online is too difficult in an adversarial setting, so algorithms for the multi-armed
bandits problem aim to instead minimize \emph{regret}: how much worse their total reward
was than the best possible total reward in hindsight. To achieve this, bandits algorithms
learn which actions have historically given more reward, and play those more frequently.
In our setting, the ``choices" are our
generator candidates, and the rewards are given by success or failure of a
generator to yield a value. Under this analogy, an algorithm for the Multi-Armed
Bandits problem will let us learn which generators give the best results
\textit{while simultaneously} generating a stream of valid inputs for the
function.

The algorithm we will use is called UCB1 \cite{auer2002finite}. While more sophisticated
bandits algorithms exist, this one is sufficient for our purposes. When given a list of generators
$g_1,\dots g_k$ and a number of rounds $T$ to run for, UCB1 runs the generators $g_i$ in a way that attempts
to learn which generator succeds the most frequently, while also attempting to not waste time by running
failing generators too much. In the end, UCB1 emits the $m \leq T$ successfully-generated values over the course of its run.
The main upshot is that the ratio success ratio $\frac{m}{T}$ of this derived generator should, in the limit,
be no worse than the \emph{best} generator the algorithm was given. 
% UCB1 achieves a
% regret bound of $O(\sqrt{KT\log K})$. While algorithms are known which achieve
% $O(\sqrt{KT})$ regret, UCB1 is simple to implement and understand, and performs
% more than well enough for our setting. In \autoref{alg:ucb1}, we present the
% UCB1 algorithm, modified for use in property-based testing. The algorithm takes
% generator candidates $g_1,\dots g_k$, the property $P$, a number of rounds to
% run for $T$, and returns a list of generated values satisfying $P$.
\begin{restatable}{thm}{regretthm}
Fix a property $P$ and generators $g_1,\dots,g_K$, where $g_i$ succeeds with probability at least $p_i$. Then,
\vspace{-2px}
\[
\lim_{T \to \infty} \mathbb{E}\left[\frac{1}{T}\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|\right] \geq \max_i p_i
\]
\end{restatable}
% Intuitively, this theorem justifies our intuition that the ucb1 algorithm
% ``finds the best generator". As the number of rounds gets large, the ratio of
% the number of valid inputs generated by \texttt{ucb1} to the number of rounds is
% no worse than the success probability of the best generator. 
In other words, the stream of values
emitted by an ``infinite run" of UCB1 acts like a
generator for $P$ which succeeds with probability at least $\max_i p_i$.

@@

I have implemented the generator inference algorithm described in this abstract,
as well as a deductive program verifier (a la Dafny or Frama-C \cite{dafny,framac}) for the
\texttt{IMP} language which uses these generators as a verification backend.
The code is available \href{https://github.com/jdublu10/triple-testing}{here}.

% Because the generator inference algorithm implemented by the prototype is the
% first of its kind, there are no existing tools we can compare it to in a
% benchmark. Instead, we simply present our results as a table
% (\autoref{tab:data}) of data about runs of the algorithm on different inputs. We
% leave it to the reader to decide if these results are impressive.
In Table~\ref{tab:data-selected}, we present some results.
In each trial, we run the generration algorithm on the specified constraint five
times for $T = 1000$ iterations. The largest issue is the performance
differences between the runs on the proposition $0 \leq x \leq y \leq 100$ with
and without explicitly including the implicand $x \leq 100$.
The ``best" generator for the proposition without it samples an
arbitrary $x \geq 0$, which is very unlikely to leave room for some $y$
satisfying $x \leq y \leq 100$. Including the added constraint ensures that we leave space
with our initial choice of $x$.

@@

\begin{table}[h]
\label{tab:data-selected}
\centering
\scalebox{0.9}{\parbox{\linewidth}{%
\begin{tabular}{|l|l|l|l|l|}
\hline
Constraint & $K$ & $R$ & $U$ \\ \hline
     $x = y + z \wedge x,y,z \geq 0$      &   $5$   &    $506$      &     $494$   \\ \hline
     $x = y + z \wedge x \geq y \wedge x,y,z \geq 0$      &   $4$  &    $48.6$      &    $951.4$    \\ \hline
     $x - y \leq 5 \wedge x - y \leq 10 \wedge y - z \leq 2$ & $3.8$ & $26.2$ & $831$ \\ \hline
     $0 \leq x \leq y \leq 100$ & $2$ & $993.4$ & $6.6$ \\ \hline
     $0 \leq x \leq y \leq 100 \wedge x \leq 100$ & $2$ & $11$ & $831$ \\ \hline
\end{tabular}
\caption{Results. $K$ = average number of unique scripts, $R$ = average number of failed draws, $U$ = average number of unique values generated.}
}}
\end{table}

@@
